Detection of slow slip events using wavelet analysis of GNSS recordings

### 2022-03-10 paper/paper_bssa.tex
Add citations to paper

\begin{itemize}
\item We use a wavelet-based signal processing method to detect transients in GNSS data, such as slow slip events.
\item There is a good correlation between detections of slow slip using GNSS data and using tremor data.
\item The method could be applied in regions where no tremor are detected in conjunction with slow slip events.
\end{itemize}

@@

Slow slip events are a new feature discovered in the last two decades in many subduction zones thanks to recordings of the displacement of Earth's surface by dense Global Navigation Satellite System (GNSS) networks. As with ordinary earthquakes, slow slip events represent slip on a fault, for instance the plate boundary between a tectonic plate subducting under another tectonic plate. However, they take a much longer time (several days to several years) to happen relative to ordinary earthquakes. They have a relatively short recurrence time (months to years) compared to the recurrence time of regular earthquakes (up to several hundreds of years), allowing scientists to observe and study many complete event cycles, which is typically not possible to explore with traditional earthquake catalogs ~\citep{BER_2011}. A slow slip event on the plate boundary is inferred to happen when there is a reversal of the direction of motion at GNSS stations, compared to the secular interseismic motion. Slow slip events have been observed in many places, such as Cascadia, Nankai (southwest Japan), Alaska, Costa Rica, Mexico, and New Zealand ~\citep{BER_2011,AUD_2016}. \\

In many places, tectonic tremor is also observed in relation to slow slip, but it is more abundant in some places. Tremor is a long (several seconds to many minutes), low amplitude seismic signal, with emergent onsets, and an absence of clear impulsive phases. Tectonic tremor have been explained as a swarm of small, low-frequency earthquakes (LFEs) ~\citep{SHE_2007_nature}, which are small magnitude earthquakes (M $\sim$ 1) for which frequency content (1-10 Hz) is lower than for ordinary earthquakes (up to 20 Hz). In subduction zones such as Nankai and Cascadia, tectonic tremor observations are spatially and temporally correlated with slow slip observations ~\citep{OBA_2002,ROG_2003}. Due to this correlation, these paired phenomena have been called Episodic Tremor and Slip (ETS). However, this is not always the case. For instance, in northern New Zealand, tremor are more challenging to detect, and seem to be located downdip of the slow slip on the plate boundary ~\citep{TOD_2016}. In Alaska, the tremor zone only partially overlaps the long-term slow slip zone and there does not appear to be any temporal correlation between tremor and slow slip occurrence ~\citep{WEC_2016}. \\

In Cascadia, there are robust signals in both GNSS and tremor. This is also the case in Nankai, where tiltmeters are used instead of GNSS. It is thus possible to use tremor as a proxy to observe slow slip events that are not directly observed in the GNSS data. For instance, ~\citet{AGU_2009} studied 23 ETS events in Cascadia with more than 50 hours of tectonic tremor. For all these events, they computed both the GPS-estimated moment release and the cumulative number of hours of tectonic tremor recorded. They observed a linear relationship between moment release and number of hours of tremor for ETS events of moment magnitude 6.3 to 6.8. Based on this linear relationship, it is possible to infer the existence of smaller slow slip events of magnitude 5-6 occurring simultaneously with smaller tremor bursts of duration 1 to 50 hours occurring in between the big ETS events, and for which there is no detectable signal in the GPS data. \\

@@

However, these methods cannot be applied to detect slow slip events in places where tremor and slow slip occurrence are not well spatially and temporary correlated, tremor is not abundant, or the seismic network is not robust enough. We thus need other methods to be able to better detect and quantify slow slip. \\

Wavelets methods such as the Discrete Wavelet Transform (DWT) are mathematical tools for analyzing time series simultaneously in the time and the frequency domain by observing how weighted differences of a time series vary from one period to the next. Wavelet methods have been widely used for geophysical applications (e.g. ~\citep{KUM_1997}). However, few studies have used wavelet methods to analyze recordings of slow slip, and their scope was limited to the detection of the bigger (magnitude 6-7) short-term (a few weeks) events ~\citep{SZE_2008,OHT_2010,WEI_2012,ALB_2019}. \\

@@

\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/tremor_longer_detail_6.pdf}
\caption{Same as Figure 6 but for the 6th level detail. The thresholds are 0.3 (for the GPS) and 0.009 (for the tremor).}
\label{pngfiguresample}
\end{figure}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/GPS_michel_detail_5.pdf}
\caption{Top: Stacked 5th level details of the wavelet decomposition of the displacement over all the GPS stations located in a 50 km radius of a given point, for the 16 red triangles indicated in Figure 3. The red lines represent the timings of the ETS events from Table 1. The blue lines represent the timings of the magnitude 5 events from the catalog of ~\citet{MIC_2019}.}
\label{pngfiguresample}
\end{figure}

>>>>

\begin{itemize}
\item We use a wavelet-based signal processing method to detect transients in GNSS data, such as slow slip events.
\item There is a good correlation between detections of slow slip using GNSS data and using tremor data.
\item The method could be applied in regions where no tremor is detected in conjunction with slow slip events.
\end{itemize}

@@

Slow slip events are a new feature discovered in the last two decades in many subduction zones thanks to recordings of the displacement of Earth's surface by dense Global Navigation Satellite System (GNSS) networks ~\citep{VER_2010,SCH_2010,JIA_2012,WAL_2012}. As with ordinary earthquakes, slow slip events represent slip on a fault, for instance the plate boundary between a tectonic plate subducting under another tectonic plate. However, they take a much longer time (several days to several years) to happen relative to ordinary earthquakes. They have a relatively short recurrence time (months to years) compared to the recurrence time of regular earthquakes (up to several hundreds of years), allowing scientists to observe and study many complete event cycles, which is typically not possible to explore with traditional earthquake catalogs ~\citep{BER_2011}. A slow slip event on the plate boundary is inferred to happen when there is a reversal of the direction of motion at GNSS stations, compared to the secular interseismic motion. Slow slip events have been observed in many places ~\citep{BER_2011,AUD_2016}, such as Cascadia ~\citep{BAR_2020}, Nankai ~\citep{NIS_2013}, Alaska ~\citep{LI_2016}, Costa Rica ~\citep{JIA_2012}, Mexico ~\citep{RAD_2012}, and New Zealand ~\citep{WAL_2020}. \\

In many places, tectonic tremor is also observed in relation to slow slip, but it is more abundant in some places ~\citep{HAL_2018}. Tremor is a long (several seconds to many minutes), low amplitude seismic signal, with emergent onsets, and an absence of clear impulsive phases. Tectonic tremor have been explained as a swarm of small, low-frequency earthquakes (LFEs) ~\citep{SHE_2007_nature}, which are small magnitude earthquakes (M $\sim$ 1) for which frequency content (1-10 Hz) is lower than for ordinary earthquakes (up to 20 Hz). In subduction zones such as Nankai and Cascadia, tectonic tremor observations are spatially and temporally correlated with slow slip observations ~\citep{OBA_2002,ROG_2003}. Due to this correlation, these paired phenomena have been called Episodic Tremor and Slip (ETS). However, this is not always the case. For instance, in northern New Zealand, tremor is more challenging to detect, and seems to be located downdip of the slow slip on the plate boundary ~\citep{TOD_2016}. In Alaska, the tremor zone only partially overlaps the long-term slow slip zone and there does not appear to be any temporal correlation between tremor and slow slip occurrence ~\citep{WEC_2016}. \\

In Cascadia, there are robust signals in both GNSS and tremor ~\citep{HAW_2013}. This is also the case in Nankai ~\citep{HIR_2008}, where tiltmeters are used instead of GNSS. It is thus possible to use tremor as a proxy to observe slow slip events that are not directly observed in the GNSS data. For instance, ~\citet{AGU_2009} studied 23 ETS events in Cascadia with more than 50 hours of tectonic tremor. For all these events, they computed both the GPS-estimated moment release and the cumulative number of hours of tectonic tremor recorded. They observed a linear relationship between moment release and number of hours of tremor for ETS events of moment magnitude 6.3 to 6.8. Based on this linear relationship, it is possible to infer the existence of smaller slow slip events of magnitude 5-6 occurring simultaneously with smaller tremor bursts of duration 1 to 50 hours occurring in between the big ETS events, and for which there is no detectable signal in the GPS data. \\

@@

However, these methods cannot be applied to detect slow slip events in places where tremor and slow slip occurrence are not well spatially and temporally correlated, tremor is not abundant, or the seismic network is not robust enough. We thus need other methods to be able to better detect and quantify slow slip. \\

Wavelet methods such as the Discrete Wavelet Transform (DWT) are mathematical tools for analyzing time series simultaneously in the time and the frequency domain by observing how weighted differences of a time series vary from one period to the next. Wavelet methods have been widely used for geophysical applications (e.g. ~\citet{KUM_1997}). However, few studies have used wavelet methods to analyze recordings of slow slip, and their scope was limited to the detection of the bigger (magnitude 6-7) short-term (a few weeks) events ~\citep{SZE_2008,OHT_2010,WEI_2012,ALB_2019}. \\

@@

\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/tremor_longer_detail_6.eps}
\caption{Same as Figure 6 but for the 6th level detail. The thresholds are 0.3 (for the GPS) and 0.009 (for the tremor).}
\label{pngfiguresample}
\end{figure}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm},clip]{figures/GPS_michel_detail_5.eps}
\caption{Top: Stacked 5th level details of the wavelet decomposition of the displacement over all the GPS stations located in a 50 km radius of a given point, for the 16 red triangles indicated in Figure 3. The red lines represent the timings of the ETS events from Table 1. The blue lines represent the timings of the magnitude 5 events from the catalog of ~\citet{MIC_2019}.}
\label{pngfiguresample}
\end{figure}
